Trauma treatment delay

### 2023-12-14 manuscriptlongversion.Rmd
text

*Introduction:* Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
<!-- Consider clarifying that delayed treatment may be one of the most 
common OFI-->
opportunities for improvement, e.g. delayed treatment. *Aim:* This
retrospective cohort study aims to identify patient level factors that
<!-- I would use "associated" rather than "linked"-->
are linked with receiving delayed treatment among trauma patients.
<!-- The first sentence is a bit complicated. Try to simplify, break 
into multiple sentences -->
*Material and Methods:* This study uses the database of Karolinska
university hospital, the equivalent of a level 1 trauma center, to run a
logistical regression linking the outcome, delayed treatment, to
variables, observed patient level factors. Patients treated for trauma
at the hospital between 2012 and 2022 were included, with ISS\>9 and/or
trauma code activation. Exclusion criterias were not being screened for
OFI and missing values in any of the variables, age, gender, ISS, SBP, RR,
GCS and number of injuries. *Results:* Higher ISS and RR was linked to
higher rates of delay in treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR,
OR 1.04, 95% CI 1.01-1.08), while a lower GCS also was linked to delayed
treatment (OR 0.93, 95% CI 0.87-1.00). Number of injuries showed a
correlation to higher rates of delay, but not when adjusted for. The
remaining variables showed no correlation. *Conclusion:* Our study
suggests that it is mostly more severely injured and less awake patients
that receive delayed treatment.

*Inledning:* Trauma är fortfarande en global hälsoutmaning som skördar
miljontals liv varje år. Den största effekten kan ses i låg till
medelinkomstländer, men i höginkomstländer har traumasystem utvecklats
för att sänka morbiditeten och mortaliteten relaterat till trauma. För
att ytterliggare förbättra dessa system hålls konferenser för att
förbättra vården patienter får och identifiera misstag som gjorts, så
kallade förbättringsmöjligheter, t.ex. försenad vård. *Syfte:* Denna
retrospektiva kohortstudie syftar till att identifiera patientfaktorer
kopplade till försenad vård bland traumapatienter. *Material och Metod:*
Denna studie använder Karolinska universitetssjukhusets databas,
motsvarande ett traumacenter på nivå 1, för att köra en logistisk
regression som kopplar utfallet, försenad vård, till variabler,
observerade patientnivåfaktorer. Patienter som behandlades för trauma på
sjukhuset mellan 2012 och 2022 inkluderades, med ISS\>9 och/eller
traumakodaktivering. Exklusionskriterier var om patient inte screenades
för ofi samt saknade värden i de använda variablerna, ålder, kön, ISS,
SBT, AF, GCS och antal skador. *Resultat:* Högre ISS och AF var kopplade
till högre grad av försenad vård (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR
1.04, 95% CI 1.01-1.08), medan ett lägre GCS också var kopplad till
försenad vård (OR 0.93, 95% CI 0.87-1.00). Antalet skador visade först
ett samband med högre förseningsfrekvens, men inte i den justerade
analysen. De återstående variablerna visade ingen korrelation.
*Slutsats*: Vår studie tyder på att det mestadels är svårare skadade och
mindre vakna patienter som får försenad behandling.

@@

## Trauma Statistics
<!-- I suggest you name this section something else, like "Trauma globally" 
or something like that. Also, divide the second paragraph into two or 
three paragraphs-->
Trauma is often divided into two subgroups, blunt force trauma and
penetrating trauma. Blunt force trauma is when an object or force
strikes the body, often causing bruising, broken bones or deep cuts.
Examples of blunt trauma could be car crashes, falls or direct blows to
the body. Penetrating trauma is when an object pierces the skin or body
and creates an open wound. Examples of penetrating trauma are gunshot
wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
groups, as blunt trauma patients tend to be more injured on arrival to
the hospital and as such, require more resources and are hospitalized
for a longer periods [@FITCH2019].

@@

## Trauma systems
<!-- Split long paragraphs -->
Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the
last three decades, the introduction of trauma systems has contributed
to lowering the incidence of preventable death. This is attributed to
the improvements in care for acute brain injuries and bleeding control.
The incidence of late death because of sepsis and multiple organ failure
has also been lowered, possibly a result of better and earlier
resuscitation [@Asensio_2008]. Functioning trauma systems saves lives in
the hospital, but their work outside of the hospital is just as
important. Data collection from injured patients, such as mechanism of
injury or mortality, are essential for creating databases that can be
used for research. In turn, that research can be used for planning
injury prevention programs that target the most common injuries in the
most efficient ways, e.g. teen drivers, children, specific occupations
etc. These injury prevention programs can be planned on trauma center,
organizational or government level[@ACS].

@@

## Mortality and morbidity (MoM) conferences
<!-- Consider renaming this section as "Opportunities for improvement" -->
Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. In Sweden, 48 out of
the 49 hospitals receiving trauma patients are connected to the national
registry SweTrau, which functions as a national database designed to
enable scientific processing with epidemiological technology
[@SweTrau_2021] This logging of information is important work, as a
cornerstone of trauma quality improvement programs is multidisciplinary
MoM conferences. The MoM conference is a meeting where different
specialities who work with trauma care sit down and discuss deaths and
complications in order to look for preventable factors. These
conferences are performed in many hospitals globally, almost everywhere
where there are formal medical specialty departments and sometimes in
smaller hospitals as well [@WHO_2009]. In Sweden, many hospitals have
mortality conferences on deceased patient cases, but only one has MoM
conferences to better their trauma care on patients who lived
[@SweTrau_2023]. The reason being that this expansion of MoM conferences
takes time and resources, which is why all hospitals don't implement
this.

@@

## Aim
<!-- Make sure the aim here is exactly the same as in the abstract. Also,
I suggest that your aim is to determine the associations between different 
patient level factors and receiving delayed treatment -->
The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

@@

<!-- You need to refer to tables and figures by numbers in the text, 
i.e. Figure 1, Table 1 and so on -->
Table 1 shows the characteristics of the group used in the logistic
regression. Most of the patients were men, 68%, with that being a bit
higher in the delayed treatment group, 74%. The mean value for the ISS
score was 10.18, with that value being higher in the outcome group, at
16.81. The mean age of the patients was 45.18 years old, with it being
around the same in the delayed treatment group at 45.50. The mean number
of injuries among the delayed treatment patients was at 6.33, which is
higher than in the overall group at 4.76. The mean SBP was 137.33, with
a standard deviation of 25.69. Mean respiratory rate was 18.59 with it
being slightly higher in the outcome population at 20.39. Lastly, mean
GCS was 13.93, being lower among patients with delayed treatment at
12.56.

@@

<!-- Use "associated" instead of correlated or linked etc-->
The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate (RR) both have
a trend that patient with lower values seem to have a lower incidence of
delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI 1.02-1.05)(RR,
OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR 1.03, 95% CI
1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the opposite
correlation, with higher GCS often having lower incidence of delayed
treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and adjusted (OR
0.93, 95% CI 0.87-1.00). Number of injuries had the correlation of
higher number of injuries having higher incidence of delayed treatment
unadjusted (OR 1.06, 95% CI 1.02-1.10), but that correlation was not
seen when adjusted for. The three remaining variables, age, gender and
SBP, showed no correlation with the observed outcome.

@@

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, while many analysing mortality. Some
similarities exist with which patient level factors predict the outcome,
with higher mortality being seen in patients with lower GCS and higher
ISS [@Mort_sam][@Mort_hef]<!-- Are these two references?-->. Increased age was also a factor linked to
mortality, which was not seen to be linked to delayed treatment in this
study. An older patient is more likely to have co-morbidities that might
affect their mortality but not increase their likelihood of receiving
delayed treatment the same way ISS and GCS does[@Gioffre2018].

@@

@@

<!-- Not so sure about this section. What do you mean that the logistic
regression assumes that the data is binomial? Do you mean that it requires 
a binary outcome? I think you discuss potential limitations well enough in 
the previous paragraphs, so you don't really need this -->
Another limitation with this study is that logistic regressions
assumes that the data is binomial, which the data used is not. The
analysis assumes a linear correlation between the dependent variable and
the independent variable, which is best analysed in a normal
distribution. However, Fischers test showed a significant difference
between the groups in 4 of the 7 variables, all of them being the ones
who showed an increased or decreased risk of delayed treatment, showing
that these results still are significant.

@@

<!-- Think about what components of your study is generalisible and to 
where. I think that your results are generalisible to  a large extent, in 
the sense that severly injured patients are at higher risk of delayed 
treatment, also in other contexts -->

#### Future studies

@@

In conclusion, based on this small population, the results suggests that
<!-- I would say "with affected consciousness" instead of "less awake" -->
it is mostly more injured and less awake trauma patients that receive
delayed treatment. However, due to the limitations of this study, the
result should only be seen as something to keep in mind when treating
patients as further studies need to be conducted on this subject.

>>>>

*Introduction:* Trauma remains a global health challenge, leading to
millions of deaths each year. The biggest impact can be seen in low to
middle income countries, but in high income countries, trauma systems
have been developed to lower the mortality and morbidity associated with
trauma. To further improve on these systems, conferences are held to
better the care patients receive and to identify errors made, called
opportunities for improvement, one of the more common being delayed
treatment. *Aim:* This retrospective cohort study aims to identify
patient level factors that are associated with receiving delayed
treatment among trauma patients.
<!-- The first sentence is a bit complicated. Try to simplify, break 
into multiple sentences --> *Material and Methods:* This study uses the
database of Karolinska university hospital, the equivalent of a level 1
trauma center, to run a logistical regression linking the outcome,
delayed treatment, to variables, observed patient level factors.
Patients treated for trauma at the hospital between 2012 and 2022 were
included, with ISS\>9 and/or trauma code activation. Exclusion criterias
were not being screened for OFI and missing values in any of the
variables, age, gender, ISS, SBP, RR, GCS and number of injuries.
*Results:* Higher ISS and RR was linked to higher rates of delay in
treatment (ISS, OR 1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI
1.01-1.08), while a lower GCS also was linked to delayed treatment (OR
0.93, 95% CI 0.87-1.00). Number of injuries showed a correlation to
higher rates of delay, but not when adjusted for. The remaining
variables showed no correlation. *Conclusion:* Our study suggests that
it is mostly more severely injured and less awake patients that receive
delayed treatment.

*Inledning:* Trauma är fortfarande en global hälsoutmaning som skördar
miljontals liv varje år. Den största effekten kan ses i låg till
medelinkomstländer, men i höginkomstländer har traumasystem utvecklats
för att sänka morbiditeten och mortaliteten relaterat till trauma. För
att ytterliggare förbättra dessa system hålls konferenser för att
förbättra vården patienter får och identifiera misstag som gjorts, så
kallade förbättringsmöjligheter, där försenad vård är en av de
vanligaste. *Syfte:* Denna retrospektiva kohortstudie syftar till att
identifiera patientfaktorer kopplade till försenad vård bland
traumapatienter. *Material och Metod:* Denna studie använder Karolinska
universitetssjukhusets databas, motsvarande ett traumacenter på nivå 1,
för att köra en logistisk regression som kopplar utfallet, försenad
vård, till variabler, observerade patientnivåfaktorer. Patienter som
behandlades för trauma på sjukhuset mellan 2012 och 2022 inkluderades,
med ISS\>9 och/eller traumakodaktivering. Exklusionskriterier var om
patient inte screenades för ofi samt saknade värden i de använda
variablerna, ålder, kön, ISS, SBT, AF, GCS och antal skador. *Resultat:*
Högre ISS och AF var kopplade till högre grad av försenad vård (ISS, OR
1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08), medan ett lägre
GCS också var kopplad till försenad vård (OR 0.93, 95% CI 0.87-1.00).
Antalet skador visade först ett samband med högre förseningsfrekvens,
men inte i den justerade analysen. De återstående variablerna visade
ingen korrelation. *Slutsats*: Vår studie tyder på att det mestadels är
svårare skadade och mindre vakna patienter som får försenad behandling.

@@

## Trauma globally

```{=html}
<!-- I suggest you name this section something else, like "Trauma globally" 
or something like that. Also, divide the second paragraph into two or 
three paragraphs-->
```
Trauma is often divided into two subgroups, blunt force trauma and
penetrating trauma. Blunt force trauma is when an object or force
strikes the body, often causing bruising, broken bones or deep cuts.
Examples of blunt trauma could be car crashes, falls or direct blows to
the body. Penetrating trauma is when an object pierces the skin or body
and creates an open wound. Examples of penetrating trauma are gunshot
wounds and stab wounds [@NIGMS_2020]. Differences exist between the two
groups, as blunt trauma patients tend to be more injured on arrival to
the hospital and as such, require more resources and are hospitalized
for a longer periods [@FITCH2019].

@@

## Trauma systems

<!-- Split long paragraphs -->

Trauma systems are infrastructures that exist to provide and optimize
care for injured patients starting with injury recognition and triage,
transport to appropriate trauma center, inpatient care and outpatient
follow-up. Beyond the clinical side, trauma systems work with outreach,
education and advocacy, data collection through registries, research,
funding, and disaster preparedness and response. A comprehensive and
functioning trauma system requires strong leadership and engagement at
the trauma center, regional and national level [@connolly_2018]. This
system is crucial to provide care for trauma patients, both in reducing
morbidity and mortality in this patient group. Earlier studies from
Sweden have shown that treating severely injured patients at a trauma
center is associated with a 41% lower adjusted 30-day mortality rate
compared to being treated at a non trauma care center due to them being
more capable of treating these patients, with potential survival benefit
increasing with higher injury severity [@Candefjord_2020]. Other studies
have shown similar results, showing that treating injured patients at
trauma centers is associated with a 15-30% decrease in mortality
[@Moran_2018; @Celso_2006; @MacKenzie_2006; @Haas_2012]. During the last
three decades, the introduction of trauma systems has contributed to
lowering the incidence of preventable death. This is attributed to the
improvements in care for acute brain injuries and bleeding control. The
incidence of late death because of sepsis and multiple organ failure has
also been lowered, possibly a result of better and earlier resuscitation
[@Asensio_2008]. Functioning trauma systems saves lives in the hospital,
but their work outside of the hospital is just as important. Data
collection from injured patients, such as mechanism of injury or
mortality, are essential for creating databases that can be used for
research. In turn, that research can be used for planning injury
prevention programs that target the most common injuries in the most
efficient ways, e.g. teen drivers, children, specific occupations etc.
These injury prevention programs can be planned on trauma center,
organizational or government level[@ACS].

@@

## Mortality and morbidity (MoM) conferences

<!-- Consider renaming this section as "Opportunities for improvement" -->

Many hospitals have a trauma registry where they log the patients
information and timeline of what happened. The extent of the registry
varies, with HIC having more complete registries. In Sweden, 48 out of
the 49 hospitals receiving trauma patients are connected to the national
registry SweTrau, which functions as a national database designed to
enable scientific processing with epidemiological technology
[@SweTrau_2021] This logging of information is important work, as a
cornerstone of trauma quality improvement programs is multidisciplinary
MoM conferences. The MoM conference is a meeting where different
specialities who work with trauma care sit down and discuss deaths and
complications in order to look for preventable factors. These
conferences are performed in many hospitals globally, almost everywhere
where there are formal medical specialty departments and sometimes in
smaller hospitals as well [@WHO_2009]. In Sweden, many hospitals have
mortality conferences on deceased patient cases, but only one has MoM
conferences to better their trauma care on patients who lived
[@SweTrau_2023]. The reason being that this expansion of MoM conferences
takes time and resources, which is why all hospitals don't implement
this.

@@

## Aim

```{=html}
<!-- Make sure the aim here is exactly the same as in the abstract. Also,
I suggest that your aim is to determine the associations between different 
patient level factors and receiving delayed treatment -->
```
The aim of this paper is to determine what different patient level
factors affect the risk of receiving delayed treatment at a trauma
center.

@@

```{=html}
<!-- You need to refer to tables and figures by numbers in the text, 
i.e. Figure 1, Table 1 and so on -->
```
Table 1 shows the characteristics of the group used in the logistic
regression. Most of the patients were men, 68%, with that being a bit
higher in the delayed treatment group, 74%. The mean value for the ISS
score was 10.18, with that value being higher in the outcome group, at
16.81. The mean age of the patients was 45.18 years old, with it being
around the same in the delayed treatment group at 45.50. The mean number
of injuries among the delayed treatment patients was at 6.33, which is
higher than in the overall group at 4.76. The mean SBP was 137.33, with
a standard deviation of 25.69. Mean respiratory rate was 18.59 with it
being slightly higher in the outcome population at 20.39. Lastly, mean
GCS was 13.93, being lower among patients with delayed treatment at
12.56.

@@

<!-- Use "associated" instead of correlated or linked etc-->

The analysis shows that some variables seem to have a correlation with
patients receiving delayed treatment. ISS and respiratory rate (RR) both
have a trend that patient with lower values seem to have a lower
incidence of delayed treatment, in both unadjusted (ISS, OR 1.04, 95% CI
1.02-1.05)(RR, OR 1.06, 95% CI 1.02-1.09) and adjusted analyses (ISS, OR
1.03, 95% CI 1.01-1.06)(RR, OR 1.04, 95% CI 1.01-1.08). GCS had the
opposite correlation, with higher GCS often having lower incidence of
delayed treatment, both unadjusted (OR 0.88, 95% CI 0.83-0.94) and
adjusted (OR 0.93, 95% CI 0.87-1.00). Number of injuries had the
correlation of higher number of injuries having higher incidence of
delayed treatment unadjusted (OR 1.06, 95% CI 1.02-1.10), but that
correlation was not seen when adjusted for. The three remaining
variables, age, gender and SBP, showed no correlation with the observed
outcome.

@@

Not many studies regarding trauma and patient level factors have had
delayed treatment as their outcome, while many analysing mortality. Some
similarities exist with which patient level factors predict the outcome,
with higher mortality being seen in patients with lower GCS and higher
ISS [@Mort_sammy][@Mort_hefny]. Increased age was also a factor linked
to mortality, which was not seen to be linked to delayed treatment in
this study. An older patient is more likely to have co-morbidities that
might affect their mortality but not increase their likelihood of
receiving delayed treatment the same way ISS and GCS does[@Gioffre2018].

@@

It is important to keep in mind that delay in treatment does not have a
set limit of time as a where under it is within the accepted timeframe
and if you go over it, it is considered delayed treatment. The audit
filters flag for a wait for CT over 30 minutes and a wait for operation
over 60 minutes, which then gets reviewed by previously mentioned two
specialized nurses. However, if a patient has a longer wait, it does not
necessarily mean that there was a delay. If it takes a longer time for a
trauma patient to receive a CT or operation, it might be because there
was more urgent matters who had to be taken care of, alternatively, the
MoM conference might deem it as the delay was justified for other
reasons. The point is that the mean waiting time for CT and operation is
higher than 30 and 60 minutes respectively when looking at the database.
Using this audit filter as the definition for delay would not mean
anything, as the majority of patients would be classified as having
waited for too long. Instead, it is the MoM conferences that decide if
the delay had any adverse effects for the patient, such as an injury
worsening during the time spent waiting without a clear motivation to
why it had to wait. It is these cases that they label it as the ofi "too
long until CT" or "Too long until operation".

@@

```{=html}
<!-- Not so sure about this section. What do you mean that the logistic
regression assumes that the data is binomial? Do you mean that it requires 
a binary outcome? I think you discuss potential limitations well enough in 
the previous paragraphs, so you don't really need this -->
```
Another limitation with this study is that logistic regressions assumes
that the data is binomial, which the data used is not. The analysis
assumes a linear correlation between the dependent variable and the
independent variable, which is best analysed in a normal distribution.
However, Fischers test showed a significant difference between the
groups in 4 of the 7 variables, all of them being the ones who showed an
increased or decreased risk of delayed treatment, showing that these
results still are significant.

@@

```{=html}
<!-- Think about what components of your study is generalisible and to 
where. I think that your results are generalisible to  a large extent, in 
the sense that severly injured patients are at higher risk of delayed 
treatment, also in other contexts -->
```
#### Future studies

@@

In conclusion, based on this small population, the results suggests that
it is mostly more injured and patients with affected consciousness that
receive delayed treatment. However, due to the limitations of this
study, the result should only be seen as something to keep in mind when
treating patients as further studies need to be conducted on this
subject.
